\begin{center}
    {\large \textbf{ACKNOWLEDGEMENTS}}
\end{center}

\thispagestyle{plain}

I owe my deepest gratitude to my advisor and chair, Dr. Zheng Mu, who responded to my cold call in 2021 and with whom I have had the utmost pleasure of conducting research since then. She has supported me to the fullest, teaching me critical thinking, sociological reasoning, research design, and academic writing while giving me the freedom to explore my own interests. She is not only an outstanding mentor but also a cherished friend. She embodies the kind of scholar, mentor, and individual I aspire to become. Without her guidance and support, none of this would have been possible.

I'm also incredibly grateful to my committee members, Prof. Bussarawan Puk Teerawichitchainan and Dr. Senhu Wang, and my examiners, Prof. Vincent Chua and Prof. Kriti Vikram, for their unwavering support and evaluation of my work. From the Qualifying Examination to the Oral Defense, they have read the manuscripts multiple times and provided me with invaluable feedback that was integral in transforming simple curiosities and ideas into academic publications. Prof. Puk has taught me extensively about social demography and the importance of meticulous data analysis and coherent argumentation, Dr. Wang has helped me sharpen my research design and methodological rigor, and Prof. Chua and Prof. Vikram have provided me with invaluable feedback on both this work and future research.

I would also like to express my heartfelt thanks to my friends in Singapore. Without Linqiu Li and Shan Yang, both incredibly kind and courageous individuals, I would not have been able to navigate the ups and downs of my personal and academic life. I am also deeply grateful to my peers Ya Guo, Jinhan Liu, Nanxun Li, and Xinyi Chen, who have been exceptional friends and colleagues. Sasha and Charlie, two community cats, have also found a special place in my heart during and beyond my time in Singapore.

Finally, I want to express my profound gratitude to those closest: my parents, who have always been my strongest supporters and greatest inspirations, Harvard, my fluffy friend who taught me love, and my beloved partner Anni Ni, whose love and companionship have transformed my life in ways words can barely capture. Together, they have made these past four years the most beautiful chapter of my life. This dissertation is as much a testament to their love and support as it is to my academic endeavors.

\clearpage